---
title: "Chronological Variograms with Simulated Process Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronological Variograms with Simulated Process Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7)
library(chRonostatistics)
```

Chronological process data should be kept in the order in which it was
collected. If the order is removed, the short-term sampling variation, longer
process movement and repeating operating cycles become mixed together.

This vignette uses three simulated examples informed by the treatment of time
series and variograms in Chapter 9 of *Statistical Methods for Mineral
Engineers*. The analysis order and interpretation of short- and long-range
variograms, variance components and engineering limits follow Part X of
Pitard's *Theory of Sampling and Sampling Practice*, Third Edition (2019).
The data used here are original simulations generated from known process
components. This makes the examples reproducible and allows the behaviour of
each variogram to be explained.

The default focused simulator is deliberately process-like: it has a modest
positive drift, small operating cycles and low measurement noise. Larger
cycles or more local variation can still be requested explicitly when they are
useful for a teaching example.

For a lag of $j$ observations, the chronological semivariogram is

$$
V(j) = \frac{1}{2N_j}\sum_{i=1}^{N_j}(x_{i+j}-x_i)^2,
$$

where $N_j$ is the number of valid pairs. `chrono_variogram()` carries out
this calculation directly from the ordered vector and retains the pair count
for every lag.

## Example 1: a long record with gradual process movement

The first example represents 360 sequential tails copper results. A low-level
autocorrelated component describes changes between adjacent samples, while two
cycles describe shorter operating adjustments and slower changes in the plant.
The seed fixes the random values, so the same series is produced each time.

```{r example-93-data}
example_93 <- simulate_chronostatistics(
  n = 360,
  interval = 1,
  mean = 0.165,
  trend = 0,
  ar = 0.80,
  innovation_sd = 0.0035,
  cycle_period = c(60, 180),
  cycle_amplitude = c(0.025, 0.010),
  measurement_sd = 0.004,
  events = FALSE,
  seed = 903
)

tails_cu <- chrono_data(
  example_93,
  time,
  value,
  interval = 1,
  units = "% Cu"
)

summary(tails_cu)
```

The first check is deliberately simple: confirm the number of observations,
segments, spacing and scale before calculating any variogram.

```{r example-93-variogram}
v93 <- chrono_variogram(tails_cu, max_lag = 180, min_pairs = 30)
ggplot2::autoplot(v93)
```

At short lags, observations are still related and $V(j)$ is relatively small.
The semivariance increases as the pairs move further apart. The rise and fall at
longer lags is caused by the simulated cycles; it should not be read as random
noise around a flat sill.

This is why lag range matters. A short variogram is useful for studying sampling
and local process variation. A longer variogram is needed before making a
judgement about a cycle or slow process movement.

```{r example-93-components}
components_93 <- chrono_components(v93, cycle_lag = 60)
components_93[, c(
  "V0", "V1", "interval_process_variance",
  "cycle_time", "cycle_contribution", "sill"
)]

ggplot2::autoplot(v93) +
  ggplot2::geom_vline(
    xintercept = components_93$cycle_time,
    linetype = "dotted"
  ) +
  ggplot2::geom_hline(
    yintercept = components_93$sill,
    linetype = "dashed"
  )
```

The cycle is selected from process knowledge and the plotted minimum. It is not
chosen automatically because two genuine cycles, an operating change or an
aliased sampling interval can produce similar shapes.

After inspecting the variogram, return to the complete chronological series to
check whether the interpretation is compatible with the observed process:

```{r example-93-timeseries}
ggplot2::autoplot(tails_cu)
```

## Example 2: a repeating operating cycle

The second example is shorter and has a much stronger repeating component. It
could represent an hourly plant measurement affected by a 24-hour operating
cycle. Its scale is deliberately similar to a plant rate rather than a grade.

```{r example-94}
example_94 <- simulate_chronostatistics(
  n = 133,
  interval = 3600,
  mean = 287,
  trend = 0,
  ar = 0.45,
  innovation_sd = 12,
  cycle_period = 24,
  cycle_amplitude = 42,
  measurement_sd = 10,
  events = FALSE,
  seed = 904
)

plant_rate <- chrono_data(
  example_94,
  time,
  value,
  interval = 3600,
  units = "t/h"
)

summary(plant_rate)
```

```{r example-94-variogram}
v94 <- chrono_variogram(plant_rate, max_lag = 65, min_pairs = 30)
ggplot2::autoplot(v94)
components_94 <- chrono_components(v94, cycle_lag = 24)
ggplot2::autoplot(v94) +
  ggplot2::geom_vline(xintercept = components_94$cycle_time, linetype = "dotted")
components_94[, c("cycle_time", "cycle_contribution", "sill")]
```

The corresponding full time series is then checked to see whether the
repeating variogram structure is visible in the operating record:

```{r example-94-timeseries}
ggplot2::autoplot(plant_rate)
```

Pairs separated by about half a cycle tend to lie on opposite sides of the
wave, so their squared differences are large. Pairs separated by one complete
cycle are in similar positions and the semivariance falls. The repeated peaks
and troughs therefore describe a time structure in the process, not a failure
of the variogram calculation.

## Example 3: short-lag extrapolation

The final example contains 133 observations with a small upward drift, local
autocorrelation and a weaker cycle. The chronological plot should be inspected
before reducing this behaviour to a variance component.

```{r example-95-data}
example_95 <- simulate_chronostatistics(
  n = 133,
  interval = 1,
  mean = 89,
  trend = 0.018,
  ar = 0.70,
  innovation_sd = 1.4,
  cycle_period = 18,
  cycle_amplitude = 2.4,
  measurement_sd = 1.3,
  events = FALSE,
  seed = 905
)

process_measurement <- chrono_data(
  example_95,
  time,
  value,
  interval = 1,
  units = "process units"
)

summary(process_measurement)
```

There is no observed pair at lag zero because an observation minus itself is
always zero. The sampling contribution is instead estimated by extrapolating
the local integral variogram towards the vertical axis. The default Pitard
method uses the first three eligible short lags and applies
the $V(0)/2$ endpoint correction. Use the modern intercept method for the
earlier AICc-selected local fitting method.

```{r example-95-variogram}
v95 <- chrono_variogram(process_measurement, max_lag = 65, min_pairs = 30)
ggplot2::autoplot(v95)
short_v95 <- v95[v95$lag <= 12, ]
components_95 <- chrono_components(v95, cycle_lag = 18)
selected <- attr(components_95, "selected_rows")
i95 <- chrono_integral_variogram(v95, integration = "pitard")
i95$W_corrected <- i95$W + components_95$V0_raw / (2 * i95$lag)

ggplot2::ggplot(
  i95[i95$lag <= 12, ],
  ggplot2::aes(lag_time, W_corrected)
) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    data = i95[selected, ],
    method = "lm",
    formula = y ~ x,
    se = FALSE
  ) +
  ggplot2::geom_hline(
    yintercept = components_95$V0,
    linetype = "dashed"
  ) +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Elapsed lag, t[j]",
    y = "Corrected Pitard first integral, W(j)"
  )

components_95[, c(
  "V0", "V1", "interval_process_variance",
  "intercept_lags", "trend"
)]
```

Finally, inspect the complete series again. This keeps the short-lag
extrapolation connected to the actual process behaviour rather than treating
it as a standalone numerical result.

```{r example-95-timeseries}
ggplot2::autoplot(process_measurement)
```

`V0` is the estimated sampling variance. The difference between `V1` and `V0`
is reported as the interval process contribution. These quantities are useful
engineering summaries, but they remain dependent on the short-lag shape and
the stability of the process.

## Reading the variance components

The variogram can be treated as a combination of variation acting over
different time scales. The short-range discontinuity represents sampling and
measurement variation. Continuous movement represents changes in the process.
Repeating peaks and troughs represent periodic behaviour. Some point-to-point
roughness also comes from estimating each ordinate with a finite number of
pairs.

These ideas are related, but the reported values should not all be added
together. Since $V(1)$ already contains $V(0)$, the interval process
contribution is $V(1)-V(0)$ rather than the full value of $V(1)$.

```{r component-definitions}
component_definitions <- data.frame(
  quantity = c(
    "V(0)",
    "V(1)",
    "V(1) - V(0)",
    "Cycle contribution",
    "Sill",
    "Trend"
  ),
  package_output = c(
    "V0 / sampling_variance",
    "V1",
    "interval_process_variance",
    "cycle_contribution",
    "sill",
    "trend"
  ),
  value = c(
    components_95$V0,
    components_95$V1,
    components_95$interval_process_variance,
    components_95$cycle_contribution,
    components_95$sill,
    NA_real_
  ),
  interpretation = c(
    "Sampling, subsampling, preparation and measurement variation extrapolated to zero lag",
    "Sampling and measurement variation plus process change over one sampling interval",
    "Process movement occurring within one sampling interval",
    "Pitard-style half amplitude with the minimum tangent anchored at V(0)",
    "Long-range variance level used as the descriptive total-variance benchmark",
    "Direction of the fitted local integral; reported as a diagnostic, not as a variance"
  ),
  stringsAsFactors = FALSE
)

component_definitions
components_95$trend
```

### A decomposed variance table

For a compact engineering summary, the sill can be used as the variance
benchmark. The table below assigns parts of that benchmark to sampling and
measurement, movement over one sampling interval, and the selected cycle. The
amount not assigned to those terms is retained as an unallocated remainder.

```{r variance-decomposition}
variance_decomposition_95 <- variance_breakdown(components_95)
knitr::kable(
  variance_decomposition_95,
  digits = c(0, 4, 4, 1, 0),
  caption = "Descriptive variance breakdown relative to the sill"
)
```

The first three rows are estimated engineering contributions. The remainder is
calculated as

$$
\text{remainder} =
\text{sill} -
\left\{V(0) + [V(1)-V(0)] + V_{\mathrm{cycle}}\right\}.
$$

It is not automatically a new physical variance component. It can contain
longer-range process movement, cycles that were not selected, estimation
differences and residual variogram roughness. A negative remainder is also
possible because the intercept, cycle and sill are estimated using different
parts of the variogram. In that case, the component estimates overlap and
should be reviewed rather than forced to reconcile.

The standard-deviation column puts each nonnegative variance contribution back
on the scale of the measured variable. The percentage column is useful for
comparing contributions within this example, but it is descriptive: it does
not turn the decomposition into an inferential model.

For a reusable report-style explanation rather than a bare data frame, call
`report(components_95)`. The returned object contains an overview, key
findings, caveats and the same `variance_table` used above. This makes the
output easier to review in an analysis report while keeping the estimates
available for custom tables and plots.

## Example 4: a synthetic copper-slag case study

Part X illustrates a copper-slag stream sampled every three hours, with a
target of 10% Cu and specifications of 7% and 14% Cu. The example below is a
new simulation at the same scale; it does not reproduce the book's data. It
has a modest short-range measurement component and a 19-sample (57-hour)
cycle.

```{r example-copper-slag}
copper <- simulate_chronostatistics(
  n = 217,
  interval = 3 * 3600,
  mean = 10,
  trend = 0,
  ar = 0.6,
  innovation_sd = 1.2,
  cycle_period = 19,
  cycle_amplitude = 1.7,
  measurement_sd = 1.1,
  events = FALSE,
  seed = 351
)

summary(copper)

copper_data <- chrono_data(
  copper, time, value,
  interval = 3 * 3600,
  units = "% Cu",
  target = 10,
  lower_spec = 7,
  upper_spec = 14
)

copper_variogram <- chrono_variogram(
  copper_data,
  max_lag = 60,
  min_pairs = 30
)
ggplot2::autoplot(copper_variogram)

copper_components <- chrono_components(
  copper_variogram,
  cycle_lag = 19,
  cycle_peak_lag = 13
)
report(copper_components)
ggplot2::autoplot(copper_variogram, components = copper_components)
```

The explicit `cycle_peak_lag` records the analyst's graphical choice. The
Pitard-style cycle contribution uses the selected peak/tangent and $V(0)$ to
estimate a half-amplitude. It is not an automatic peak detector. Return to the
full chronological record before deciding whether the inferred cycle has a
credible operating explanation:

```{r example-copper-timeseries}
ggplot2::autoplot(copper_data)
```

## Example 5: a synthetic paper-scale iron-ore stream

Minnitt and Pitard's iron-ore example uses stratified assays collected every
four hours. It describes a short cycle around 17 lags and a larger cycle around
208 lags, or about 2.8 and 35 days. The simulation below uses those engineering
scales and a similar mean and standard deviation, but all observations are
newly generated for this vignette.

```{r example-iron-ore}
iron <- simulate_chronostatistics(
  n = 1511,
  interval = 4 * 3600,
  mean = 66.1,
  ar = 0.75,
  innovation_sd = 0.18,
  cycle_period = c(17, 208),
  cycle_amplitude = c(0.28, 0.50),
  measurement_sd = 0.25,
  events = FALSE,
  seed = 2008
)

iron_data <- chrono_data(
  iron, time, value,
  interval = 4 * 3600,
  units = "% Fe",
  target = 66.1,
  lower_spec = 65,
  upper_spec = 67.5
)

iron_variogram <- chrono_variogram(
  iron_data, max_lag = 300, min_pairs = 30
)
iron_components <- chrono_components(
  iron_variogram, cycle_lag = 208
)

plot(iron_data)
ggplot2::autoplot(iron_variogram, components = iron_components)
iron_components[, c("V0", "V1", "cycle_contribution", "sill")]
report(iron_components)
```

Supplying `components` adds a Pitard-style interpretation layer to the
variogram: horizontal lines mark $V(0)$, $V(1)$ and the sill, while the
selected `cycle_lag` is marked with the estimated cyclic contribution. The
cycle remains user-defined because a variogram peak alone cannot distinguish
equipment, feed, maintenance and sampling-alias effects.

The paper's main practical lesson is that the variogram makes time-scale
structure visible. The short cycle should not automatically be treated as
random noise, and the long cycle should not be removed without understanding
whether it comes from feed, equipment, maintenance or operator practice.

## Example 6: an operating plant table

The larger simulator produces multiple related variables rather than one
isolated series. Online product quality can be analysed at the hourly grid,
while laboratory results retain their own sparse reporting process.

```{r example-plant-process}
plant <- simulate_plant_process(n = 24 * 90, seed = 20260717)

product <- chrono_data(
  plant,
  timestamp,
  product_fe,
  interval = 3600,
  units = "% Fe"
)

plant_variogram <- chrono_variogram(
  product, max_lag = 24 * 14, min_pairs = 30
)

summary(product)
plot(product)
plot(plant_variogram)
```

Rows where the plant is unavailable become segment boundaries. This prevents a
lag pair from silently spanning a shutdown. The `lab_result_available` column
is intentionally not passed directly to `chrono_data()` as an hourly series:
it represents delayed, sparse laboratory reporting and should be regularised
according to the analytical question first.

## Optional tsibble workflow

Users who prefer tidyverse syntax can keep their data in a tibble and pass
named columns directly to `chrono_data()`. If the data are already stored as a
`tsibble`, the index can be inferred and the measured variable supplied
directly. This section is optional: `tsibble`, `fable`, `feasts` and
`fabletools` are not required for the core chronological analysis.

```{r example-tibble, eval=FALSE}
plant_tbl <- dplyr::transmute(
  plant,
  timestamp,
  product_fe,
  availability
)

product <- chrono_data(
  plant_tbl,
  time = timestamp,
  value = product_fe,
  interval = 3600
)
```

With a `tsibble` installed:

```{r example-tsibble, eval=FALSE}
plant_ts <- tsibble::as_tsibble(
  plant,
  index = timestamp,
  key = ore_domain
)
product <- chrono_data(plant_ts, value = product_fe, interval = 3600)
```

## Integral conventions

The default `pitard` method retains the discrete cumulative construction used
for the Part X endpoint correction. The `trapezoid` option integrates against
physical lag from $V(0)=0$ and is useful as a transparent comparison. Keeping
the choice explicit allows the two numerical conventions to be compared for a
new analysis.

Write $j$ for the discrete lag index and $t_j$ for the corresponding elapsed
lag. For the first integral, write

$$
W(j) = \frac{1}{j}\int_0^j V(u)\,du.
$$

This is an average of the semivariogram from zero lag to lag $j$; it does not
create new observations. With discrete lags, the package approximates the area
between successive ordinates. The Part X endpoint correction then writes the
working value as $W(j) = W_0(j) + V(0)/(2j)$, where $W_0(j)$ is the discrete
curve before the unknown zero-lag endpoint is restored. Regressing these local
values against $t_j$ and solving that endpoint relation gives the extrapolated
$V(0)$. The integral smooths noisy short-lag ordinates so this extrapolation is
less dependent on any one pair estimate.


```{r integration-comparison}
trapezoid <- chrono_integral_variogram(v95, integration = "trapezoid")
pitard <- chrono_integral_variogram(v95, integration = "pitard")

integration_comparison <- rbind(
  transform(as.data.frame(trapezoid), method = "trapezoid"),
  transform(as.data.frame(pitard), method = "pitard")
)

ggplot2::ggplot(
  integration_comparison[integration_comparison$lag <= 20, ],
  ggplot2::aes(lag_time, W, colour = method)
) +
  ggplot2::geom_line() +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "Lag", y = "First integral variogram, W(j)")
```

## Engineering limits

Progressive limits show how the estimated sampling, interval and cyclic
contributions widen an engineering operating band. The linear-root convention
and a root-sum-square comparison are both available. Neither should be described
as a probability-based prediction interval without a separate statistical
model that supports that interpretation.

In the Pitard linear-root presentation, the interval term is $V(1)$. In the
RSS comparison it is the non-overlapping interval-process contribution
$V(1)-V(0)$. Each output row records the exact quantities included in its
width.

```{r limits}
linear_limits <- chrono_control_limits(components_95)
rss_limits <- chrono_control_limits(components_95, convention = "rss")

rbind(
  as.data.frame(linear_limits),
  as.data.frame(rss_limits)
)[, c("stage", "lower", "upper", "convention")]

plot(linear_limits)
```

## Checks before using plant data

The timestamps must be ordered and regularly spaced. Missing observations and
large time gaps create separate segments, and lag pairs do not cross those
segments by default. Smaller departures from the declared spacing are rejected
because an index lag would then combine different elapsed times.

The chronological plot and the operating history should be reviewed with the
variogram. Ore-domain changes, shutdowns, maintenance, changing targets and
sensor faults can all create structure that resembles a cycle or trend.
Chronostatistics helps separate variation by time scale, but it does not replace
the process knowledge needed to explain that variation.

## References

Napier-Munn, T. J. (2025). *Statistical Methods for Mineral Engineers: How to
Design Experiments and Analyse Data*, revised edition, Chapter 9. Julius
Kruttschnitt Mineral Research Centre, The University of Queensland.
[Book information from JKTech](https://jktech.com.au/statsbook).

Pitard, F. F. (2019). *Theory of Sampling and Sampling Practice*, third
edition, Part X: Chronostatistics. CRC Press.
[Publisher information and DOI](https://doi.org/10.1201/9781351105934).

Minnitt, R. C. A., and Pitard, F. F. (2008). *Application of variography to
the control of species in material process streams: %Fe in an iron ore
product*. The Journal of the Southern African Institute of Mining and
Metallurgy, 108, 109–122. The example scales are adapted as a synthetic
teaching scenario; the paper's observations and figures are not redistributed.
